Is Our Community Getting Safer in Recent Years: A Data-Driven Study of Crime Victims in Toronto

### 2024-09-24 paper/paper.qmd
retouch again, add one more table

Our findings show that while most crime types, such as assault, have exhibited a consistent decresing trend in the last decade, they all seem to regain the upward trends in or after 2020, posting new threats and challenges to public safety where youth and child victimization trends are of particular concern. These findings contribute to ongoing public safety discussions in Toronto, emphasizing the need for targeted crime prevention strategies and improved public policy.

In the body of this paper, we will first discuss the data source and our computational environment, along with some basic facts and significances about the data in @sec-data. We then present the analysis results,  focusing on crime subtype trends, gender distribution, and the impact of recent social challenges in @sec-results to address our primary research question for the trend of public safety in recent years in the City of Toronto. Finally in @sec-discuss, we conclude the paper with a summary of key findings and their implications for public safety in Toronto and discuss existing limitations and future directions.

@@

The dataset used in this study, "Police Annual Statistical Report: Victims of Crime," is provided by the City of Toronto through its open data initiative. [1] It reflects crime incident reports, focusing on the experiences of victims from 2019 to 2023. The data is publicly accessible, ensuring transparency in city governance and fostering community engagement in crime monitoring. Ethically, the data adheres to the Open Government License, which mandates privacy protections for individuals involved, promoting responsible use in academic and policy analysis. [2] Statistically, the dataset offers insights into crime trends, aiding in evidence-based policymaking. The data set also provides rich context for statisticians to extract useful insights by analyzing victim components and trends over time.

@@

The dataset has no missing values in each of its column, and we will not be analyzing the column `ASSAULT_SUBTYPE` because limited information is found in the Open Data Toronto data description page. [1] Hence no further data cleaning is needed other than removing the `ASSAULT_SUBTYPE` column. This comprehensive dataset facilitates focused analysis on victim profiles and crime trends, making it valuable for understanding community safety.

@@

We first attempt to discover trends in crime over time across different subtypes and identify specific areas that might require targeted interventions. We first consider @fig-one below which focuses on analyzing on crime dynamics in Toronto from 2014 to 2023. Assault has a prominent increasing trend, starting around $15000$ victims in 2014 and rising to over $20000$ by 2023, indicating a substantial escalation in assaults over the past decade except the sudden drop in 2019. On the other hand, crimes categorized as "Other" display a gradual decline after a peak in 2015, stabilizing in recent years, but this subtype is also witnessing a reclimb starting from 2022. Robbery and sexual violations show generally more stable, moderate patterns with minor fluctuations.

@@

As we look closer into the largest crime subtype of assaults in @fig-two, we discover that majority of victims are adults, regardless of gender, with over $70000$ adult victims in both male and female categories. It is worth noting that the victims of unknown age group are at comparative figures as child and youth victims, also posting concerns for the data quality being recorded in the police system. Additionally, while youth and child victims appear to be fewer, they still require further attentions especially after we adjust for the population age distribution.

@@

As we discovered in @sec-results-1, all crime subtypes seem to regain the growth trend starting in or after 2020. This demands a closer examination of the dynamics within the victim data. Table 2 below reveals that the gender distribution of crime victims has remained relatively balanced over the years. This stability in gender distribution suggests a consistent pattern in the way crimes affect male and female populations. More importantly, the two genders both demonstrate a strictly increasing trend in the total number of victims with a growth rate of over $10\%$ from 2021 to 2022. This observation could be evidence for the residual effects of the COVID-19 pandemic and the accompanying economic challenges. The pandemic's disruptions could have contributed to a rise in both opportunity-driven and stress-induced crimes, placing general public mental health condition at another challenge.

@@

As we breakdown the age groups into finer ranges, @fig-three shows that there are consistently a number of outliers in the adult victim category for each year, suggesting that while most adult incidents fall within a consistent range, certain periods saw spikes in crime involving adults, possibly due to heightened vulnerability during specific events. This illustrates the needs for stricter regulations on public events.

Moreover, a slight upward trend in the number of youth and adult victims is observed after 2020, which is likely connected to broader societal impacts, such as increased stress and economic instability following the COVID-19 pandemic. This shift may indicate rising safety concerns faced by the younger populations over time, and calls for a more robust satefy awareness education strategy.

@@

In this study, we start from introducing the open-source dataset that describes crime victims in Toronto in the last ten years, and then proceed with figure and table visualizations to address the research question of whether our community is getting safer in recent years. From our analysis in @sec-results, we discover that assault cases represent the largest proportion of crimes, showing consistent growth over the years. The gender distribution of victims remains balanced, with no significant differences between male and female victimization rates.

However, crime rates across all subtypes began to rise again after 2020 as seen by the increasing numbers of victims across all subtypes, age groups, and genders. The number of youth and child victims also demonstrate upward trends in the new decade with about $75\%$ increase in the number of youth victims which raises particular concerns. Even though recent improvements in data collection methods show positive progress in identifying victims more accurately, public safety is still facing an increasing number of challenges. Based on the data-driven results, we are centainly not confident enough to claim significant progress for the public safety in the Toronto communities.

@@

There are still a few limitations that should be acknowledged in this study. The dataset offers a broad overview without including more specific details on the impacts or contexts of individual crimes, limiting our ability to analyze causes and consequences of crime activities in greater depth. Additionally, the analysis was based on visualization without incorporating advanced statistical models, such as regression analysis or time series forecasting, to better quantify trends or correlations. For instance, cross-factor interactions—like how age and sex may jointly affect crime victimization—remain unexplored.

>>>>

Our findings show that while most crime types, such as assault, have exhibited a consistent decreasing trend in the last decade, they all seem to regain the upward trends in or after 2020, posing new threats and challenges to public safety, where youth and child victimization trends are of particular concern. These findings contribute to ongoing public safety discussions in Toronto, emphasizing the need for targeted crime prevention strategies and improved public policy.

In the body of this paper, we will first discuss the data source and our computational environment, along with some basic facts and significance about the data in @sec-data. We then present the analysis results, focusing on crime subtype trends, gender distribution, and the impact of recent social challenges in @sec-results to address our primary research question for the trend of public safety in recent years in the City of Toronto. Finally, in @sec-discuss, we conclude the paper with a summary of key findings and their implications for public safety in Toronto and discuss existing limitations and future directions.

@@

The dataset used in this study, "Police Annual Statistical Report: Victims of Crime," is provided by the City of Toronto through its open data initiative. [1] It reflects crime incident reports, focusing on the experiences of victims from 2019 to 2023. The data is publicly accessible, ensuring transparency in city governance and fostering community engagement in crime monitoring. Ethically, the data adheres to the Open Government License, which mandates privacy protections for individuals involved, promoting responsible use in academic and policy analysis. [2] Statistically, the dataset offers insights into crime trends, aiding in evidence-based policymaking. The dataset also provides rich context for statisticians to extract useful insights by analyzing victim components and trends over time.

@@

The dataset has no missing values in any of its columns, and we will not be analyzing the column `ASSAULT_SUBTYPE` because limited information is found in the Open Data Toronto data description page. [1] Therefore, no further data cleaning is needed other than removing the `ASSAULT_SUBTYPE` column. This comprehensive dataset facilitates focused analysis on victim profiles and crime trends, making it valuable for understanding community safety.

@@

We first attempt to discover trends in crime over time across different subtypes and identify specific areas that might require targeted interventions. We first consider @fig-one below which focuses on analyzing crime dynamics in Toronto from 2014 to 2023. Assault has a prominent increasing trend, starting around $15000$ victims in 2014 and rising to over $20000$ by 2023, indicating a substantial escalation in assaults over the past decade except for the sudden drop in 2019. On the other hand, crimes categorized as "Other" display a gradual decline after a peak in 2015, stabilizing in recent years, but this subtype is also witnessing a resurgence starting from 2022. Robbery and sexual violations show generally more stable, moderate patterns with minor fluctuations.

@@

As we look closer into the largest crime subtype of assaults in @fig-two, we discover that the majority of victims are adults, regardless of gender, with over $70000$ adult victims in both male and female categories. It is worth noting that the victims of unknown age group are at similar figures to child and youth victims, also raising concerns about the data quality being recorded in the police system. Additionally, while youth and child victims appear to be fewer, they still require further attention especially after we adjust for the population age distribution.

@@

As we discovered in @sec-results-1, all crime subtypes seem to regain the growth trend starting in or after 2020. This demands a closer examination of the dynamics within the victim data. Table 2 below reveals that the gender distribution of crime victims has remained relatively balanced over the years. This stability in gender distribution suggests a consistent pattern in the way crimes affect male and female populations. More importantly, both genders demonstrate a strictly increasing trend in the total number of victims, with a growth rate of over $10%$ from 2021 to 2022. This observation could be evidence for the residual effects of the COVID-19 pandemic and the accompanying economic challenges. The pandemic's disruptions could have contributed to a rise in both opportunity-driven and stress-induced crimes, placing the general public's mental health in further jeopardy.

@@

As we break down the age groups into finer ranges, @fig-three shows that there are consistently a number of outliers in the adult victim category for each year, suggesting that while most adult incidents fall within a consistent range, certain periods saw spikes in crime involving adults, possibly due to heightened vulnerability during specific events. This illustrates the need for stricter regulations on public events.

Moreover, a slight upward trend in the number of youth and adult victims is observed after 2020, which is likely connected to broader societal impacts, such as increased stress and economic instability following the COVID-19 pandemic. This shift may indicate rising safety concerns faced by the younger populations over time, and calls for a more robust safety awareness education strategy.

@@

In this study, we start by introducing the open-source dataset that describes crime victims in Toronto in the last ten years, and then proceed with visualizing figures and tables to address the research question of whether our community is getting safer in recent years. From our analysis in @sec-results, we discover that assault cases represent the largest proportion of crimes, showing consistent growth over the years. The gender distribution of victims remains balanced, with no significant differences between male and female victimization rates.

However, crime rates across all subtypes began to rise again after 2020 as seen by the increasing numbers of victims across all subtypes, age groups, and genders. The number of youth and child victims also demonstrates upward trends in the new decade, with about a 20% increase in the number of adult victims, and a 75% increase in youth, which raises particular concerns. Even though recent improvements in data collection methods show positive progress in identifying victims more accurately, public safety is still facing an increasing number of challenges. Based on the data-driven results, we are certainly not confident enough to claim significant progress in public safety in the Toronto communities.

@@

There are still a few limitations that should be acknowledged in this study. The dataset offers a broad overview without including more specific details on the impacts or contexts of individual crimes, limiting our ability to analyze the causes and consequences of criminal activities in greater depth. Additionally, the analysis was based on visualization without incorporating advanced statistical models, such as regression analyses or time series forecasts, to better quantify trends or correlations. For instance, cross-factor interactions, such as how age and sex may jointly affect crime victimization, remain unexplored.
